---
title: "Writing and maintaining the Zcalc blog"
description: |
  Instructions to authors and maintainers of this blog.
author:
  - name: Daniel Kaplan
date: 2022-01-27
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

When writing for this blog, it's important to keep in mind the overall structure.

The home page for the blog is contained in `index.Rmd` at the top level of the repository. That document, when compiled to HTML, will **not** generate a listing of blog entries. Instead, it is an ordinary `{distill}` Rmd file with links to the various blog directories.

Those blog directories are `_Block-1`, `_Block-2`, ..., `_Block-6`, `_Manifestations`, and `_Instructions`. When creating a new posting, be sure to place it in the correct directory. For instance, to create a new posting relevant to Block-5 of the course, create the skeleton of the post this way:

```{r eval=FALSE}
distill::create_post("My new post", "Block-5")
```

The second argument is the directive to put the new post on the `Block-5` page.

Along with the various `_Block-{n}` directories in the top level of the repo, there are files named `Block-1.Rmd`, `Block-2.Rmd`, and so on. These appear to have no content, but actually the content will be created when the website is built. After being built, each of those documents becomes a listing of the posts in the corresponding `_Block-{n}` directory. *You should not need to edit the `Block-{n}.Rmd` files.*


