---
title: "142Z Project 2: Dynamics"
author: "Daniel Kaplan"
---

```{r include=FALSE}
library(etude2)
library(mosaic)
library(mosaicCalc)
```

## Objectives

MAKE THE PROJECT ABOUT THE DYNAMICS OF NEWTON'S METHOD, Maybe about zero-finding and optimization on functions of two variables.


## Raw material that was previously an exercise.

Last semester we introduced an *iterative* technique for finding the zeros of a function $f(x)$, that is, finding a value $x^\star$ such that $f(x^\star) = 0$. The method, called Newton's method, involves making a starting guess $x_0$ for the location of the zero and then refining this guess according to the famous formula $$x_{i+1} = x_i - \frac{f(x_i)}{\partial_x f(x_i)}$$

We can use this method to find, for instance, $\sqrt{10}$. We do this by creating a function that's easy to calculate which has a zero at $x^\star = \sqrt{10}$:
$$f(x) \equiv x^2 - 10$$

Finding $\partial_x f(x)$ is easy, so the famous formula becomes, for the special case of our $f()$
$$x_{i+1} \equiv N(x_i) \equiv x_i - \frac{x_i^2 - 10}{2 x_i}$$
We've called the dynamics function $N()$, in honor of Newton. Applying $N()$ to $x_i$ is called "taking a Newton step."

Start with a very rough guess for $x^\star$, say $x_0 = 1$. Applying the formula once gives $x_1 = 1 - -9/2 = 5.5$. 

Using the sandbox, implement the function $N(x)$. Apply it once to $x_0=1$ to make sure that you have the dynamics right. 

```{r nstep, exercise=TRUE, exercise.cap="Newton steps", exercise.nlines=6}
N <- makeFun( 0 ~ x)
x_0 <- 1
N(x_0)    # this is x_1
N(N(x_0)) # this is x_2
```
```{r nstep1, echo=FALSE}
etude2::etudeQ(
  "Using $N()$ as the dynamics and starting with $x_0 = 1$, what is $x_5$?",
  "5.5" = "That's $x_1$.",
  "3.659091" = "That's $x_2$.",
  "3.141593" = "That's $\\pi$, which incidentally is $\\neq \\sqrt{10}$.",
  "3.196005" = "That's $x_3$.",
  "3.162456" = "That's $x_4$.",
  "3.162354",
  "+3.162278+",
  random_answer_order = FALSE
)
```

```{r nstep2, echo=FALSE}
etude2::etudeQ(
  "Modify `N()` to find $\\sqrt{20}$. Starting at $x_0=1$, how many times do you have to apply your new `N()` to get an answer right to within 1% of the true number?",
  "After 2 steps we get 4.202" = "No, after 2 steps you would get 6.202, which is about 40% away from the true answer.",
  "After 3 steps we get 4.713." = "That's about 6% away from the true number.",
  "After 3 steps we get 4.472." = "That's not the right answer for $x_3$.",
  "After 4 steps we get 4.472." = "That's not the right answer for $x_4$.",
  "+After 4 steps we get 4.478.+" = "Right. A bit closer than 1% to the true answer.",
  random_answer_order = FALSE
)
```

```{r nstep3, echo=FALSE}
etude2::etudeQ(
  "Modify your `N()` once again to find $\\sqrt[3]{10}$. (That is, the cube-root of 10.) Starting at $x_0 = 1$, take 3 Newton steps. What is $x_3$?",
  "2.154" = "That's $x_{5}$.",
  "+2.320+",
  "2.875" = "That's $x_2$.",
  "2.912",
  random_answer_order = FALSE
  
)
```
