---
title: Instructor notes: Assembling functions (Block 1, 2nd chap, new version)
---


## Classroom examples of piecewise functions

1. As a classroom activity, show students the graph of the ramp function and ask them to define it, using both traditional and computer notation.

```{r echo=FALSE}
#| column: margin
#| fig-pos: "h"
ramp <- makeFun(ifelse(0 < x, x, 0) ~ x)
slice_plot(ramp(x) ~ x, domain(x=c(-3, 3)))
```

The ***ramp function*** is a kind of one-sided absolute value:
$$\text{ramp}(x) \equiv \left\{
\begin{array}{cl} x & \text{for}\ 0 \leq x\\0 & \text{otherwise}\end{array}
\right.$$
Or, in computer notation:

```{r results="hide"}
ramp <- makeFun(ifelse(0 < x, x, 0) ~ x)
slice_plot(ramp(x) ~ x, domain(x=c(-3, 3)))
```

2. Ask students to use three-pieces of the domain to define this chunky. From easier to harder:

i. identify the three pieces
2. write down the leftmost and rightmost functions
3. construct the linear function for the middle piece
4. put them together in mathematical notation. 
5. implement the function in R. You can do this by nesting one `ifelse()` inside another 

```{r}
sig <- 
  makeFun(
    ifelse(x < -1, 
           0, # leftmost branch
           ifelse(x > 1, 1, (x+1)/2)) # rightmost and middle branches
    ~ x)
slice_plot(sig(x) ~ x, domain(x=c(-3, 3)), npts=501)
```  


A more elegant construction uses a linear combination of two input-shifted ramp functions.

```{r}
sig <- makeFun(0.5*(ramp(x+1) - ramp(x-1)) ~ x)
```
